Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-05 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Updated regression table analysis and conclusion

The research question aims to measure the COVID-19 case rate per 100,000 residents within each state based on the make-up of the population, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable variables, and to the level at which the effect of these variables differ help. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation

@@

Following this, in the second phase of investigation the analysis considers other variables of state-level population that are considered factors of contractability, namely the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018) and the rate of unemployment (defined as the percent unemployment in 2018). Those living below the poverty line may have less access/are unable to afford to preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home as well as access to the aforementioned preventative controls. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states.

@@

In model 1, the coefficient of -0.174 for the senior rate per 100,000 is significant with a p-value of 0.0234 (per the regression results from the Model 1 section). When the covariate variables of unemployment rate and poverty rate are added in to model 2, the coefficient for senior rate adjusts to a greater magnitude of -0.231 while the p-value appreciably decreases to a value of 0.0015 (per the regression results from the Model 2 section). This explains that not only has the magnitude of effect of the senior rate on case rate increased, but that this increased effect is also more statistically significant. In addition, the coefficients of the covariate variables themselves are also highly significant with p-values of 0.0005 and 0.0059 for poverty rate and unemployment rate respectively. However, it should be noted that the magnitudes of these effects are very small, with poverty rate having a positive relationship of 0.002 per 100,000 with case rate while unemployment rate has a negative relationship of 0.005 per 100,000.

Moving onto model 3, the magnitude of the coefficient for senior rate increases once again to a value of -0.242 while the statistical significance increases with a decreased p-value of 0.0004 (per the regression results from the Model 3 section). As highlighted before, the more complex model does a better job of predicting the effect of the senior rate variable. Conversely, the p-values for the poverty rate and unemployment rate coefficients become less significant, as they increase to 0.002 and 0.049 respectively. However, there is almost no detectable change in the coefficient values, as poverty rate is still at 0.002 and unemployment rate increases to -0.004. The change in the p-values are likely a direct result of the added variables of mask policy and the logarithmic transform of population density, where these variables introduce some additional variance in the model 2 covariates. Notably, the coefficient for the mask policy covariate is significant with a p-value of 0.0027 and has by far the greatest effect on the outcome variable with a coefficient of -945.6. The populuation density covariate is not significant with a p-value of 0.3369.

In terms of practical significance, there is great discrepancy between the effect of the mask policy variable compared to the other three significant features. With a coefficient of -945.6, this means that given the binary mask policy variable, when a mask policy is implemented (value of 1), there is an decrease in case rate of roughly 945 per 100,000 compared to when the policy is not in place. The next largest effect is from the senior rate variable, where in the final model, the coefficient of -0.242 translates as a decrease of 0.242 cases per 100,000 with every additional senior per 100,000. More intuitively, it can be said that there is an decrease of roughly 1 case per 100,000 with every 4 additional seniors per 100,000. This effect is roughly 190 times smaller than the effect of wearing a mask. Similarly, the effects of both poverty rate and unemployment rate are nominal. For every 500 additional people per 100,000 who are below the poverty line, there is 1 additional case per 100,000, and for every 250 additional people per 100,000 who are unemployed, there is 1 fewer case per 100,000. In comparison with likewise minimal effect of the senior rate, these effects are 100 and 50 times smaller respectively.

@@

In conclusion, although there is no evidence to show that although there is a direct correlation between the COVID case rate and population density as originally hypothesized in our research question, it can be seen from further development of the models that there is some correlation in other variables examined, namely the mask mandate variable. As can be seen from model 3, the regression model gives a mask mandate variable coefficient with p-value of 0.003058, which is highly significant. The negative coefficient also suggests that the original conjecture that the implementation of mask policies within a state would decrease the COVID case rate was indeed accurate.

Conversely, the hypotheses that population and demographics would have an effect on the COVID case rate was unable to be accepted. Although upon investigating the poverty percentage variable in model 2 there was a detection of notable adjacency to a significant result (and an actual significant result when using the `stargazer` package to compute the regression table) with a p-value of 0.05755, it was not enough to be truly significant and reject the null hypothesis. Similarly, the relationship of race/ethnicity as well as age on case rate was also insignificant with no apparent correlation.

Given these results in conjunction with the original aim of the research question, it can be suggested that variables that can be controlled (in this case, mask policy) has a greater effect on stymieing the spread of COVID-19 than the variables that may be inherent to a state, such as population density and age/ethnicity demographics. This is a promising insight, as it provides opportunity for states to implement and apply policies to intervene and control the case rate without being worried about its varying effects depending on population characteristics. However, it should once again be stressed that the final report may include and analyze alternative variables and apply them to different models based on the findings from this draft report, which as a result may also lead to more concrete or slightly varied insights.

>>>>

The research question aims to measure the COVID-19 case rate per 100,000 residents within each state based on the make-up of the population, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable variables, and to the level at which the effect of these variables differ help. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation.

@@

Following this, in the second phase of investigation the analysis considers other variables of state-level population that are considered factors of suscesptibility, namely the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018) and the rate of unemployment (defined as the percent unemployment in 2018). Those living below the poverty line may have less access/are unable to afford to preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home as well as access to the aforementioned preventative controls. Alternatively, the likelihood of reduced mobility for those living under the poverty line and/or who are unemployed may also play a role in the relationship with case rate. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states.

@@

In model 1, the coefficient of -0.174 for the senior rate per 100,000 is significant with a p-value of 0.0234 (per the regression results from the Model 1 section). When the covariate variables of unemployment rate and poverty rate are added in to model 2, the coefficient for senior rate adjusts to a greater magnitude of -0.231 while the p-value appreciably decreases to a value of 0.0015 (per the regression results from the Model 2 section). This explains that not only has the magnitude of effect of the senior rate on case rate increased, but that this increased effect is also more statistically significant. In addition, the coefficients of the covariate variables themselves are also highly significant with p-values of 0.0005 and 0.0059 for poverty rate and unemployment rate respectively. The magnitude of the effect of poverty rate is almost equivalent to that of the senior rate, with a single increase in poverty rate per 100,000 equivalent to 0.232 per 100,000 for case rate, while the unemployment rate has stronger negative coefficient of -0.486 per 100,000 (roughly twice the magnitude of the other two variables).

Moving onto model 3, the magnitude of the coefficient for senior rate increases once again to a value of -0.242 while the statistical significance increases with a decreased p-value of 0.0004 (per the regression results from the Model 3 section). As highlighted before, the more complex model does a better job of predicting the effect of the senior rate variable. Conversely, the p-values for the poverty rate and unemployment rate coefficients become less significant, as they increase to 0.002 and 0.049 respectively. There are also detectable changes in the coefficient values, as the magnitude for both decreases to 0.197 and -0.364 respectively. The changes in the p-values are likely a direct result of the added variables of mask policy and the logarithmic transform of population density, where these variables introduce some additional variance in the model 2 covariates. Notably, the coefficient for the mask policy covariate is significant with a p-value of 0.0027 and has by far the greatest effect on the outcome variable with a coefficient of -945.6. The populuation density covariate is not significant with a p-value of 0.3369.

In terms of practical significance, there is great discrepancy between the effect of the mask policy variable compared to the other three significant features. With a coefficient of -945.6, this means that given the binary mask policy variable, when a mask policy is implemented (value of 1), there is an decrease in case rate of roughly 945 per 100,000 compared to when the policy is not in place. The next largest effect is from the unemployment rate variable, where in the final model, the coefficient of -0.364 translates as a decrease of 0.364 cases per 100,000 for every additional increase of 1 in unemployment rate per 100,00. More intuitively, it can be said that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This effect is roughly 2,600 times smaller than the effect of implementing a mask policy. Similarly, the effects of both poverty rate and senior rate are nominal. For every 5 additional people per 100,000 who are below the poverty line, there is roughly 1 additional case per 100,000, and for every 4 additional seniors per 100,000 within the state population, there is roughly 1 fewer case per 100,000. In comparison with likewise relatively minimal effect of the unemployment rate, these absolute effects on case rate are approximately one-third (1/3) and one-half (1/2) smaller respectively.

@@

The goal of the analysis was to address the research question to examine the relationship between both the distribution of population demographics and mask-related policy of a state and the COVID-19 case rate within that states. In order to address this question, the analysis was broken up into three phases of investigation with corresponding linear models:
- Phase 1: To investigate the direct relationship between COVID-19 case rate and the population distribution of seniors within a state
- Phase 2: To build upon Phase 1 by including two additional variables of population demographics, poverty rate and unemployment rate, and investigating the overall effect on COVID-19 case rate
- Phase 3: To build upon Phase 2 by adding a variable to measure the effect of the implementation of a mask-policy on COVID-19 case rate. A variable for population density was also added as an additional measure of population demographics

From the three phases of investigation, there was strong evidence in the data to suggest that senior rate, unemployment rate, and mask policy are all statistically significant in relation to case rate and have a negative relationship with case rate, while poverty rate has a positive relationship. The data for population density was not statistically significant enough to suggest a relationship. Out of the four significant variables, the greatest effect was from the mask policy variable, followed by the unemployment rate, and then similar absolute effect sizes from the senior rate and povery rate.

Starting with the variable with the greatest effect size, it can be seen that the implementation of a mask policy in a state leads to a decrease of 945 cases per 100,000. Given that research from leading health organizations shows that the wearing of masks helps to decrease the chance of the transmission of COVID-19, it could be suggested that a mask policy is effective in increasing the adoption of wearing masks among the state population. Next, it can be seen that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This suggests that as unemployment increases, the case rate decreases, which could be a result of the fact that those who are unemployed may be less mobile as they are not commuting into work, lowering the likelihood of contracting the virus. Additionally, at the time that this data was collected, the Coronavirus Aid, Relief, and Economic Security (CARES) Act was still supporting the unemployed with relief payments so there was less urgency to be out looking for work. Following unemployment, when examining the poverty variable there is actually an increase of roughly 1 case per 100,000 for every 5 additional people per 100,000 living under the federal poverty line. Although there is some intuition behind the unemployment and poverty rates having similar effects on case rate, it can be seen from the test of collinearity in the Model 2 section that these two variables are not explaining the same effects. This difference is likely due to the fact that those living below the poverty line have less access to suitable living conditions and sanitary goods that help to alleviate the possibility of COVID-19 infection, while those who were recently employed up until the pandemic hit may not be in the same economical situation. Finally, the effect of the senior rate is negative, with an additional 4 seniors per 100,000 translating to roughly 1 fewer case per 100,000. As touched upon in the introduction, this could be largely down to the fact that it is widely broadcasted that seniors fall within the high risk category in terms of having severe complications from COVID-19, and therefore they are more diligent in the precautions they take against contracting the virus.

In conclusion, we see that although the population demographics had varying effects on the COVID-19 case rate, the greatest effect was seen from the mask policy. Going back to our original question, it can be suggested within the data that the effect of a mask-related policy is much larger than that of population demographics. Therefore, it seems that the COVID-19 case rate at the state-level is much more dependent on masks, and that ultimately, there are effective policy and control measures that can be taken by state governments that have a greater contribution towards combating it's spread relative to factors that cannot be controlled, such as population demographics. However, it should be noted that these insights are qualified by the assumptions and limitations laid out throughout the report, namely that there are potential issues in meeting the IID assumption and that there is explanatory information that is lost from both the variables investigated as well as the those that were not included in the dataset.
